Experiences in Modelling a Microcontroller Instruction Set Using B

### 2012-01-19 trunk/papers/ArticleZ80/source/paper.tex
Abstract updated, (Please, David look the new abstract).

  This paper describes an approach to model the functional aspects of
  the instruction set of microcontrollers and several details about
  the representation of microcontrollers.  Different models were
  developed using the notation of the B method. This paper presents
  specifically a B-based formal model of the Z80 instruction set.

%  \paragraph[Keywords]{
%   \textbf{Keywords}: Embedded Software, Simulation, B method and Verification%}
\end{abstract}

@@

This work presents a formal modell of the instruction set of the Z80 
microcontroller~\cite{Z80_manual}. This microcontroller
was chosen because of several factors: it contains essential and common
concepts in microcontrollers and until microprocessors; it has extensive
documentation available; and it was widely used and is still in use. 
A first Z80 modell presented in \cite{Valerio_SBMF09} was verified completely,
but it uses constructs that are not well supported to animate. 
This modelling was changed to support animation in ProB\cite{proB}.
Basically, the new modelling\footnote{The interested reader in more
details is invited to visit our repository at: http://code.google.com/p/b2asm/.}
changed elements represented by infinite sets and adjusted the implications
of this change. Using the responsibility division mechanisms provided by B, auxiliary
libraries of basic modules were developed as part of the construction of microcontroller
model. Such library has many definitions about common concepts used in
the microcontrollers;besides the new Z80 model, a theoretic case study in 
petroleum production test system was developed using Z80 assembly language
to analyse the code simulation and verification process.

The formal model of instruction set provides many benefits. The animation of 
Z80 model allows to simulate the execution assembly programs including
support for interrupts and input and output instructions. Other possible uses
of a formal model of a microcontroller instruction set include documentation,
the construction of simulators, and can be possibly the starting point of a 
verification effort for the actual implementation of a Z80 design. Moreover, 
the model of the instruction set could be instrumented with non-functional
aspects, such as the number of cycles it takes to execute an instruction, to
prove lower and upper bounds on the execution time of a routine.
Two aspects of these formalization are of particular importance to us:
provide a more solid documentation artifact for microcontrollers, 
build a reference model for a formal compilation approach in B.

\subsection{Problems in formalizing instruction set}
% Compact version 
In general, the manuals of microcontrollers can be more suited for
developers in assembly language. Many times, each instruction of microcontrollers
is shown in its official manual using textual description, math description, examples,
encoding, number of cycles and other information. Some times, the instruction descriptions
have an own notation, some informations are organized in different sections of document 
and the textual description are not standardized. This does not allow the interpretation by parser.
The manuals also have semi-formal and informal descriptions, these descriptions may make mistakes.
The official manual of the microcontroller Z80 has also several problems that were identified
over time~\cite{UndocumentedZ80}.

@@

This paper is structured as it follows. %Section~\ref{sec:B_method} provides a short introduction to the B method. 
Section~\ref{sec:models} presents the elementary libraries and
the modelling of some elements common to microcontrollers.  Section~\ref{sec:z80}
presents the B model of the Z80 instruction set. %Section ~\ref{sec:Verification} provides some information on the proof effort needed to analyse the presented models.
%Section ~\ref{sec:studycase} explains the building and verifying assembly-level refinements. 
Related works are discussed in section~\ref{sec:relatedworks}. Finally, the last section is devoted to the conclusions.
%
%\section{Introduction to the B Method}
%\label{sec:B_method}
%
%The B method for software development~\cite{Abrial} is based on the B Abstract
%Machine Notation (AMN) for the formal specification of functional requirements and the use of formally proved refinements up to a
%specification sufficiently concrete that programming code can be automatically
%generated from it. Its mathematical basis consists of first order logic, integer
%arithmetic and set theory, and its corresponding constructs are similar to those
%of the Z notation~\cite{Spivey96}.
%
%A B specification is structured in modules. A module defines a set of valid
%states, including a set of initial states, and operations that may provoke a
%transition among states. The design process starts with a module with a
%so-called functional model of the system under development. In this initial
%modelling stage, the B method requires that the user proves that, in a machine,
%all its initial states are valid, and operations do not define
%transitions from valid states to invalid states.
%
%
%Essentially, a B module contains two main parts: a header and the available
%operations. Figure~\ref{fig:maqB} has a very basic example. The clause
%\MACHINE  has the name of module.  The next two clauses, respectively,
%reference external modules and create an instance of an external module.  The
%\VARIABLES clause declares the name of the variables that compose the
%state of the machine. Next, the \INVARIANT clause defines the valid states,
%specifying  the type and other restrictions on the variables. The \INITIALISATION specifies
%the initial states. Finally, operations correspond to the transitions among
%states of the machine.
%
%
%\begin{figure}[h]
%  % \begin{small}
%\begin{center}
%\begin{sloppypar}
%
%  $\begin{array}{lcl}
%    \begin{array}[t]{l}
%      \MACHINE \quad  \textit{micro} \\
%      \SEES \quad \textit{TYPES}, \textit{ALU} \\
%      \INCLUDES  \quad \textit{MEMORY} \\
%      \VARIABLES \quad    \textit{pc}  \\
%      \INVARIANT\\ 
%      \quad \textit{pc} \in \textit{INSTRUCTION} 
%    \end{array}
%    & \hspace*{0cm} &
%    \begin{array}[t]{l}
%      \INITIALISATION  \textit{pc} := 0 \\
%      \OPERATIONS\\
%      \textit{JMP} \textit{( jump )} = \\
%      \quad \PRE \textit{jump} \in \textit{INSTRUCTION}\\ 
%      \quad \THEN \textit{pc} := \textit{jump}\\  
%      \quad \END \\
%      \END
%    \end{array}
%  \end{array}
%  $
% \end{sloppypar} 
%\end{center}
%\caption{A very basic B machine.}
%\label{fig:maqB}
%\end{figure}

@@

\subsection{Modelling registers}

@@

\hspace*{0.40in}\it zero \rm = \it is\_zero \rm ( \it io\_ports \rm ( \it rgs8 \rm ( \it c0 \rm ) \rm ) \rm )  $\land$

\hspace*{0.40in}\it half\_carry \rm = \rm 0  $\land$

\hspace*{0.40in}\it pv \rm = \it parity\_even \rm ( \it io\_ports \rm ( \it rgs8 \rm ( \it c0 \rm ) \rm ) \rm ) $\land$

\hspace*{0.40in}\it add\_sub \rm =\hspace*{0.10in}\rm 0  $\land$

\hspace*{0.40in}\it carry \rm = \it z\_c

@@

% 
%There are in the literature several
%approaches~\cite{BHDL_2003,DBLP:journals/entcs/EvansG08,DBLP:conf/fmco/Leuschel08,DBLP:conf/asm/Wright08}
%to model hardware and the virtual machines using B. In these works, the B
%has been used successfully to model the operational semantic. However the cost of modelling is still expensive. 
%
%These recent works \cite{Marc20113,springerlink:Yuan2011} have a similar approach using Event B. 
%However, our work use the B method, that seems more appropriated to software development, because it has
%implementable language defined, called B0, and tools to convert the models to a programming language.
%
%
%The main motivation of our research is the development of verified
%software up to the assembly level, which requires specifying the
%semantics of the underlying hardware. Thus some aspects were not
%modelled in our work such as the execution time of the instructions.
%Also we did not consider the micro architecture of the hardware as the scope of
%our work does not include hardware verification.

@@

% Um pequeno resumo do trabalho
This work has shown an approach to the formal modelling of the instruction set of
microcontroller using the B method. During the construction of this model, some
problems were found in the official reference for Z80
microcontroller~\cite{Z80_manual}. The designer fixed the found problems in
documentation and also developed a case study: a verified embedded software. This
case study was interesting to analyse the technique of formal verification up to
assembly level.

% TODO: Adaptar para apenas fundamentar melhor o trabalho já desenvolvido. Citar o artigo SBMF e a dissertação. 
% Os meus trabalhos já desenvolvidos/ resultados atuais.
The next works quoted are directly related to the objective this paper.
The first work \cite{DantasSemish2008} shows the developed metodology to verify
software up to assembly level using B. A second work\cite{Dantas_SBMF08}
presents more details about the verification approach and a small example
software verified up to assembly level in three different platforms. 
The last work \cite{Valerio_SBMF09} shows a general view of the first version of Z80
model and techniques used in verification process.

%Link com o parágrafo anterior 
These works created a important step to software verification up to assembly
language, they show the actuals difficulties and suggest improvements in the
B tools.

@@

\subsubsection{Acknowledgements:} This experiment could not have succeeded without the interesting work of Michael Leuschel who developed several improvements in the ProB and provided important aid to enable the animation of the Z80 modelling.

>>>>

The rigorous development of an embedded software in a microcontroller requires 
a specification of its instruction set and the software requirements specification. 
The actual industrial practice is to report the instruction set informally using natural language
or with added semi-formal notation. This practice may leave space for errors and inconsistencies \cite{springerlink:Yuan2011}.
This paper presents specifically a B formal model of the Z80 instruction set. 
This model  can be  used on documentation, simulation, test and  a model-driven approach
to design and develop software from the functional specification level down to assembly.
%Using this model-driven approach and the instruction set model, a small case study to
%analyse a petroleum production test system in oil field was developed.
%  
% \paragraph[Keywords]{ \textbf{Keywords}: Embedded Software, Simulation, B
% method and Verification%}
\end{abstract}

@@

This work presents a formal model of the instruction set of the Z80
microcontroller~\cite{Z80_manual}. This microcontroller was chosen
because of several factors: it contains essential and common concepts
of microcontrollers and even microprocessors; it has extensive
documentation available; and it has been widely used.  A first,
partial, Z80 model presented in \cite{Valerio_SBMF09} was fully
verified, but it uses constructs that make it impossible to animate.

The new model supports animation in ProB\cite{proB}.  Essentially, the
new model\footnote{The interested reader in more details is invited to
  visit our repository at: http://code.google.com/p/b2asm/.}  no
longer contains definitions derived from infinite sets. Using B
constructs for modularity, auxiliary libraries of basic modules were
developed as part of the construction of microcontroller model. Such
library has many definitions of concepts common in microcontrollers.
Besides the new Z80 model, a petroleum production test system was
developed using Z80 assembly language to analyse the code simulation
and verification process.

Providing the formal model of an instruction set yields many
benefits. The animation of Z80 model allows to simulate the execution
assembly programs including support for interrupts and input and
output instructions. Other possible uses of a formal model of a
microcontroller instruction set include documentation, the
construction of simulators, and can be possibly the starting point of
a verification effort for the actual implementation of a Z80
design. Moreover, the model of the instruction set could be
instrumented with non-functional aspects, such as the number of cycles
it takes to execute an instruction, to prove lower and upper bounds on
the execution time of a routine.  Two aspects of these formalization
are of particular importance to us: provide a more solid documentation
artifact for microcontrollers, build a reference model for a formal
compilation approach in B.

\subsection{Problems in formalizing instruction set}
% Compact version

In general, the manuals of microcontrollers can be more suited for \marginpar{DD: o título da sub-seção não condiz com seu conteúdo. Não cita nenhum problema na formalização. Não vejo muito bem o que ele traz de novo que já não foi dito anteriormente.}
developers in assembly language. Many times, each instruction of
microcontrollers is shown in its official manual using textual
description, math description, examples, encoding, number of cycles
and other information. Sometimes, the instruction descriptions have
an own notation, some informations are organized in different sections
of document and the textual description are not standardized. This
does not allow the interpretation by parser.  The manuals also have
semi-formal and informal descriptions, these descriptions may make
mistakes.  The official manual of the microcontroller Z80 has also
several problems that were identified over
time~\cite{UndocumentedZ80}.

@@

This paper is structured as it follows.  Section~\ref{sec:models}
presents the elementary libraries and the modeling of some elements
common to microcontrollers.  Section~\ref{sec:z80} presents the B
model of the Z80 instruction set.  Related works are discussed in
section~\ref{sec:relatedworks}. Finally, the last section is devoted
to the conclusions.

%
%\section{Introduction to the B Method}
%\label{sec:B_method}
%
%The B method for software development~\cite{Abrial} is based on the B Abstract
%Machine Notation (AMN) for the formal specification of functional requirements and the use of formally proved refinements up to a
%specification sufficiently concrete that programming code can be automatically
%generated from it. Its mathematical basis consists of first order logic, integer
%arithmetic and set theory, and its corresponding constructs are similar to those
%of the Z notation~\cite{Spivey96}.
%
%A B specification is structured in modules. A module defines a set of valid
%states, including a set of initial states, and operations that may provoke a
%transition among states. The design process starts with a module with a
%so-called functional model of the system under development. In this initial
%modelling stage, the B method requires that the user proves that, in a machine,
%all its initial states are valid, and operations do not define
%transitions from valid states to invalid states.
%
%
%Essentially, a B module contains two main parts: a header and the available
%operations. Figure~\ref{fig:maqB} has a very basic example. The clause
%\MACHINE  has the name of module.  The next two clauses, respectively,
%reference external modules and create an instance of an external module.  The
%\VARIABLES clause declares the name of the variables that compose the
%state of the machine. Next, the \INVARIANT clause defines the valid states,
%specifying  the type and other restrictions on the variables. The \INITIALISATION specifies
%the initial states. Finally, operations correspond to the transitions among
%states of the machine.
%
%
%\begin{figure}[h]
%  % \begin{small}
%\begin{center}
%\begin{sloppypar}
%
%  $\begin{array}{lcl}
%    \begin{array}[t]{l}
%      \MACHINE \quad  \textit{micro} \\
%      \SEES \quad \textit{TYPES}, \textit{ALU} \\
%      \INCLUDES  \quad \textit{MEMORY} \\
%      \VARIABLES \quad    \textit{pc}  \\
%      \INVARIANT\\ 
%      \quad \textit{pc} \in \textit{INSTRUCTION} 
%    \end{array}
%    & \hspace*{0cm} &
%    \begin{array}[t]{l}
%      \INITIALISATION  \textit{pc} := 0 \\
%      \OPERATIONS\\
%      \textit{JMP} \textit{( jump )} = \\
%      \quad \PRE \textit{jump} \in \textit{INSTRUCTION}\\ 
%      \quad \THEN \textit{pc} := \textit{jump}\\  
%      \quad \END \\
%      \END
%    \end{array}
%  \end{array}
%  $
% \end{sloppypar} 
%\end{center}
%\caption{A very basic B machine.}
%\label{fig:maqB}
%\end{figure}

@@

\subsection{Modeling registers}

@@

\hspace*{0.40in}\it zero \rm = \it is\_zero \rm ( \it io\_ports \rm ( \it rgs8 \rm ( \it c0 \rm ) \rm ) \rm )  $\land$ \it half\_carry \rm = \rm 0  $\land$

\hspace*{0.40in}\it pv \rm = \it parity\_even \rm ( \it io\_ports \rm ( \it rgs8 \rm ( \it c0 \rm ) \rm ) \rm ) $\land$
\it add\_sub \rm =\hspace*{0.10in}\rm 0  $\land$ 
\it carry \rm = \it z\_c

@@

% 
%There are in the literature several
%approaches~\cite{BHDL_2003,DBLP:journals/entcs/EvansG08,DBLP:conf/fmco/Leuschel08,DBLP:conf/asm/Wright08}
%to model hardware and the virtual machines using B. In these works, the B
%has been used successfully to model the operational semantic. However the cost of modeling is still expensive. 
%
%These recent works \cite{Marc20113,springerlink:Yuan2011} have a similar approach using Event B. 
%However, our work use the B method, that seems more appropriated to software development, because it has
%implementable language defined, called B0, and tools to convert the models to a programming language.
%
%
%The main motivation of our research is the development of verified
%software up to the assembly level, which requires specifying the
%semantics of the underlying hardware. Thus some aspects were not
%modeled in our work such as the execution time of the instructions.
%Also we did not consider the micro architecture of the hardware as the scope of
%our work does not include hardware verification.

@@

% Um pequeno resumo do trabalho
This work has shown an approach to the formal modeling of the instruction set of
microcontroller using the B method. During the construction of this model, some
problems were found in the official reference for Z80
microcontroller~\cite{Z80_manual}. The designer fixed the found problems in
documentation and also developed a case study: a verified embedded software. This
case study was interesting to analyse the technique of formal verification up to
assembly level.

% TODO: Adaptar para apenas fundamentar melhor o trabalho já desenvolvido. Citar o artigo SBMF e a dissertação. 
% Os meus trabalhos já desenvolvidos/ resultados atuais.
The next works quoted are directly related to the objective this paper.
The first work \cite{DantasSemish2008} shows the developed metodology to verify
software up to assembly level using B. A second work\cite{Dantas_SBMF08}
presents more details about the verification approach and a small example
software verified up to assembly level in three different platforms. 
The last work \cite{Valerio_SBMF09} shows a general view of the first version of Z80
model and techniques used in verification process.
%Link com o parágrafo anterior 
These works created a important step to software verification up to assembly
language, they show the actuals difficulties and suggest improvements in the
B tools.

@@

\subsubsection{Acknowledgements:} This experiment could not have succeeded without the interesting work of Michael Leuschel who developed several improvements in the ProB and provided important aid to enable the animation of the Z80 modeling.
